mirror determines the far posterior domain in butterfly wings

Version 1

Abstract

Abstract

Insect wings, a key innovation that contributed to the explosive diversification of insects, are recognized for their remarkable variation and many splendid adaptations. Classical morphological work subdivides insect wings into several distinct domains along the antero-posterior (AP) axis, each of which can evolve relatively independently. There has been little molecular evidence, however, for AP subdivision beyond a single compartment boundary described from Drosophila melanogaster. Here we show that the transcription factor mirror acts as a selector gene to differentiate a far posterior domain in the butterfly wing, classically defined as the vannus, and has wide-ranging effects on wing shape, scale morphology, and color pattern. Our results confirm that insect wings can have more than one posterior developmental domain, and support models of how selector genes may facilitate evolutionarily individuation of distinct AP domains in insect wings. Our results also suggest that the alula, a small mirror-dependent structure at the base of the D. melanogaster wing, may be an evolutionary derivative of the vannus, and therefore that the D. melanogaster wing blade is a solitary remigium that represents only a fraction of the archetypal insect wing.

Results and discussion

Insect wings vary dramatically in their size and morphology. Butterflies and moths in particular have attracted attention for their rapidly evolving wings that show extreme variation in shape and color pattern. Interestingly, lepidopteran color pattern diversity appears to follow some simple rules where basic spot and stripe color pattern elements change size, color, and presence/absence relatively independently based on which domain along the antero-posterior (AP) axis they occur in.3 Analyses of wing pattern diversity across butterflies, considering both natural variation and genetic mutants, suggest that wings can be subdivided into at least five AP domains, bounded by the M1, M3, Cu2, and 2A veins respectively, within each of which there are strong correlations in color pattern variation and wing morphology (Figure 1A).4-6 The anterior-most of these domains, bordered by the M1 vein, appears to correspond to an AP compartment boundary originally described by cell lineage tracing in Drosophila melanogaster,7 and later supported in butterfly wings by expression of the Engrailed transcription factor.8 Interestingly, however, D. melanogaster work has yet to reveal clear evidence for additional AP domain boundaries in the wing. It has thus been a matter of significant interest and controversy whether additional AP wing domains might occur in other insects, 9-11 as their existence would provide an explanatory model for the modular diversification of insect wing morphology.

Proposed AP domains of insect wings.

(A) Comparative morphology has identified boundaries associated with color pattern variation demarcated by the M1, M3, Cu2, and 2A veins, as summarized in McKenna et al.5 (B) Examples of the 2A vein marking a clear posterior color pattern boundary across multiple butterfly families. (C) Proximal posterior features of the archetypal dipteran wing include the calypters, which are only found in Calypterata clade, and the alula, a lobe at the base of the wing blade.1 (D) mirror RNAi knockdowns result in highly specific loss of the alula in D. melanogaster.2 (E) Snodgrass’ model of the archetypal insect wing specifies three major domains along the anteroposterior axis: the remigium, the vannus, and the jugum. Homologies of these domains with the features of the dipteran wing (C), or butterfly wing pattern boundaries (A), has remained unclear.

The goal of this study was to identify the molecular basis of AP domain specification in the butterfly wing, posterior of the Engrailed boundary. We initially identified the transcription factor mirror as a potential candidate gene based on previous mRNA-seq studies,12 coupled with D. melanogaster work showing that mirror specifies the alula – a small membranous lobe at the posterior base of the wing in some dipterans (Figure 1C,D).2,13 We used in situ hybridization to visualize mirror mRNA localization in the last-instar imaginal discs of the common buckeye butterfly, Junonia coenia, and observed mirror expression throughout the wing domain posterior to the 2A vein boundary (Figures 2A and S1). This expression precisely marks the vannus, or anal region (Figure 1E), of the butterfly wing – a distinct posterior field of the wing blade populated by the anal (A) veins, and bordered anteriorly by the claval fold between the cubital (Cu) and A veins in the hindwing. This region was also previously proposed to represent the posterior-most color pattern domain in butterflies (Figure 2B).3

mirror determines the identity of the far posterior vannus domain in J. coenia.

(A) In J. coenia mirror (red) is expressed posterior to the 2A vein in the late last-instar hindwing imaginal discs, marking the region that will develop into the vannus. See Figure S1 for forewing expression and controls. (B) In adult butterfly wings the vannus is the wing field posterior to the 2A vein, as shown here in J. coenia. (C) Targeted mosaic knockouts (mKOs) of mirror results in loss of vannal identity in J. coenia, including (D) reduction of the anal fold, and (E) continuation of margin color patterns along the posterior wing margin. Some mirror mKOs (C-E) also show mutation in the 2A vein along with loss (D) or reduction (E) of the 3A vein, likely associated with mosaic clone boundaries. mKOs result in asymmetric effects observed in the wings where (D,E) one wing resembles a wild-type wing (WT*) more than the other wing from the respective individual. For more mirror mKOs see Figures S2 and S3.

To functionally assess the role of mirror in wing development, we produced CRISPR/Cas9 deletion mosaic knockouts (mKOs) targeting the coding region of J. coenia mirror. The resulting mutants revealed vannus-specific phenotypes, consistent with the mirror expression patterns. The unique identity of the vannus was lost in mutants, so that the vannus resembled the remigium (Figure 2C-E; Figure S2). In brief, the vannus took on a morphology consistent with anterior wing domains, including loss or reduction of the vannal fold (along with an overall shape along the posterior wing margin), reduction of light silvery iridescent coloration, and continuation of normally truncated wing margin color patterns along the posterior distal margin of the wing (Figure 2E; Figure S2). Some mutants also displayed aberrations in vannal venation, including loss of the 3A vein, discontinuous and bifurcated 2A veins, and ectopic veins (Figures 2C-E; Figures S2 and S3). Together, our expression and knockout data show that mirror acts as a selector gene necessary for defining the identity of the far posterior wing domain that corresponds to the vannus.

Our finding is of interest for several reasons. First, it provides a molecular explanation for how different AP domains of the butterfly wing may be individuated to independently evolve their own shape and color pattern variations. Previous authors have proposed the existence of such individuated domains, and speculated that they may be specified by selector genes.5,10 Our data provide experimental support for this model, and now motivate us to identify factors that specify other domain boundaries between the M1 and A2 veins. Second, the role of mirror in specifying the vannus provides an important link between molecular developmental biology and Snodgrass’ classical anatomical designations of the insect wing fields – i.e., the remigium, the vannus, and the jugum (Figure 1E).14 Importantly, mirror knockdown in the milkweed bug Oncopeltus fasciatus causes developmental defects in the claval and anal furrows in the posterior wing,15 which leads us to infer that mirror’s role in determining the vannus may be deeply conserved in insects.

Finally, the function of mirror in determining the alula in D. melanogaster2 suggests that the alula may represent an evolutionarily reduced vannus. The ramifications of this are significant for reconstructing the history of the insect wing, as they suggest that the D. melanogaster wing blade is a lone remigium – only one-third of the archetypal insect wing (Figure 1E). The dipteran jugum appears to have been lost entirely, perhaps except as a calypter in the Calypterata lineage, as speculated by Snodgrass.14 Thus, while flies have been an important model for characterizing genes and processes that build insect wings, a more comprehensive understanding of the development and evolution of insect wings will require work in species that have wings more representative of the complete ancestral blueprint.

eLife Assessment

This article is a valuable addition to the growing literature on the developmental patterning of insect wings. Using CRISPR mutagenesis and localization of mRNA, the authors present solid evidence that the transcription factor Mirror is necessary for specifying the morphological identity of the most posterior regions of butterfly wings. The manuscript would benefit from more careful use of terminology and appropriate citation of related Drosophila literature, and there are also some concerns about whether the phenotype represents transformation or loss which might be clarified through a closer look at ultrastructure. With a clearer presentation of terminology, this paper would be of general interest to developmental and evolutionary biologists.

Reviewer #1 (Public Review):

Summary:

This short report shows that the transcription factor gene mirror is specifically expressed in the posterior region of the butterfly wing imaginal disk, and uses CRISPR mosaic knock-outs to show it is necessary to specify the morphological features (scales, veins, and surface) of this area.

Strengths:

The data and figures support the conclusions. The article is swiftly written and makes an interesting evolutionary comparison to the function of this gene in Drosophila. Based on the data presented, it can now be established that mirror likely has a similar selector function for posterior-wing identity in a plethora of insects.

Weaknesses:

This first version has minor terminological issues regarding the use of the terms "domains" and "compartment".

Reviewer #2 (Public Review):

This is a short and unpretentious paper. It is an interesting area and therefore, although much of this area of research was pioneered in flies, extending basic findings to butterflies would be worthwhile. Indeed, there is an intriguing observation but it is technically flawed and these flaws are serious.

The authors show that mirror is expressed at the back of the wing in butterflies (as in flies). They present some evidence that is required for the proper development of the back of the wing in butterflies (a region dubbed the vannus by the ancient guru Snodgrass). But there are problems with that evidence. First, concerning the method, using CRISP they treat embryos and the expectation is that the mirror gene will be damaged in groups of cell lineages, giving a mosaic animal in which some lines of cells are normal for mirror and others are not. We do not know where the clones or patches of cells that are defective for mirror are because they are not marked. Also, we do not know what part of the wing is wild type and what part is mutant for mirror. When the mirror mutant cells colonise the back of the wing and that butterfly survives (many butterflies fail to develop), the back of the wing is altered in some selected butterflies. This raises a second problem: we do not know whether the rear of the wing is missing or transformed. From the images, the appearance of the back of the wing is clearly different from the wild type, but is that due to transformation or not? And then I believe we need to know specifically what the difference is between the rear of the wing and the main part. What we see is a silvery look at the back that is not present in the main part, is it the structure of the scales? We are not told. There are other problems. Mirror is only part of a group of genes in flies and in flies both iroquois and mirror are needed to make the back of the wing, the alula (Kehl et al). What is known about iro expression in butterflies?

In flies, mirror regulates a late and local expression of dpp that seems to be responsible for making the alula. What happens in butterflies? Would a study of the expression of Dpp in wildtype and mirror compromised wings be useful?

Thus, I find the paper to be disappointing for a general journal as it does little more than claim what was discovered in Drosophila is at least partly true in butterflies. Also, it fails to explain what the authors mean by "wing domains" and "domain specification". They are not alone, butterfly workers, in general, appear vague about these concepts, their vagueness allowing too much loose thinking.

Since these matters are at the heart of the purpose and meaning of the work reported here, we readers need a paper containing more critical thought and information. I would like to have a better and more logical introduction and discussion.

The authors do define what they mean by the vannus of the wing. In flies the definition of compartments is clear and abundantly demonstrated, with gene expression and requirement being limited precisely to sets of cells that display lineage boundaries. It is true that domains of gene expression in flies, for example of the iroquois complex, which includes mirror, can only be related to patterns with difficulty. Some recap of what is known plus the opinion of the authors on how they interpret papers on possible lineage domains in butterflies might also be useful as the reader, is no wiser about what the authors might mean at the end of it!

The references are sometimes inappropriate. The discovery of the AP compartments should not be referred to Guillen et al 1995, but to Morata and Lawrence 1975. Proofreading is required.

Reviewer #3 (Public Review):

Summary:

The manuscript by Chatterjee et al. examines the role of the mirror locus in patterning butterfly wings. The authors examine the pattern of mirror expression in the common buckeye butterfly, Junonia coenia, and then employ CRISPR mutagenesis to generate mosaic butterflies carrying clones of mirror mutant cells. They find that mirror is expressed in a well-defined posterior sector of final-instar wing discs from both hindwings and forewings and that CRISPR-injected larvae display a loss of adult wing structures presumably derived from the mirror expressing region of hindwing primordium (the case for forewings is a bit less clear since the mirror domain is narrower than in the hindwing, but there also do seem to be some anomalies in posterior regions of forewings in adults derived from CRISPR injected larvae). The authors conclude that the wings of these butterflies have at least three different fundamental wing compartments, the mirror domain, a posterior domain defined by engrailed expression, and an anterior domain expressing neither mirror nor engrailed. They speculate that this most posterior compartment has been reduced to a rudiment in Drosophila and thus has not been adequately recognized as such a primary regional specialization.

Critique:

This is a very straightforward study and the experimental results presented support the key claims that mirror is expressed in a restricted posterior section of the wing primordium and that mosaic wings from CRISPR-injected larvae display loss of adult wing structures presumably derived from cells expressing mirror (or at least nearby). The major issue I have with this paper is the strong interpretation of these findings that lead the authors to conclude that mirror is acting as a high-level gene akin to engrailed in defining a separate extreme posterior wing compartment. To place this claim in context, it is important in my view to consider what is known about engrailed, for which there is ample evidence to support the claim that this gene does play a very ancestral and conserved function in defining posterior compartments of all body segments (including the wing) across arthropods.

(1) Engrailed is expressed in a broad posterior domain with a sharp anterior border in all segments of virtually all arthropods examined (broad use of a very good pan-species anti-En antibody makes this case very strong).

(2) In Drosophila, marked clones of wing cells (generated during larval stages) strictly obey a straight anterior-posterior border indicating that cells in these two domains do not normally intermix, thus, supporting the claim that a clear A/P lineage compartment exists.

In my opinion, mirror does not seem to be in the same category of regulator as engrailed for the following reasons:

(1) There is no evidence that I am aware of, either from the current experiments, or others that the mirror expression domain corresponds to a clonal lineage compartment. It is also unclear from the data shown in this study whether engrailed is co-expressed with mirror in the posterior-most cells of J. coenia wing discs. If so, it does not seem justified to infer that mirror acts as an independent determinant of the region of the wing where it is expressed.

(2) Mirror is not only expressed in a posterior region of the wing in flies but also in the ventral region of the eye. In Drosophila, mirror mutants not only lack the alula (derived approximately from cells where mirror is expressed), but also lack tissue derived from the ventral region of the eye disc (although this ventral tissue loss phenotype may extend beyond the cells expressing mirror).

In summary, it seems most reasonable to me to think of mirror as a transcription factor that provides important development information for a diverse set of cells in which it can be expressed (posterior wing cells and ventral eye cells) but not that it acts as a high-level regulator as engrailed.

Recommendation:

While the data provided in this succinct study are solid and interesting, it is not clear to me that these findings support the major claim that mirror defines an extreme posterior compartment akin to that specified by engrailed. Minimally, the authors should address the points outlined above in their discussion section and greatly tone down their conclusion regarding mirror being a conserved selector-like gene dedicated to establishing posterior-most fates of the wing. They also should cite and discuss the original study in Drosophila describing the mirror expression pattern in the embryo and eye and the corresponding eye phenotype of mirror mutants: McNeill et al., Genes & Dev. 1997. 11: 1073-1082; doi:10.1101/gad.11.8.1073.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public Review):

Summary:

This short report shows that the transcription factor gene mirror is specifically expressed in the posterior region of the butterfly wing imaginal disk, and uses CRISPR mosaic knock-outs to show it is necessary to specify the morphological features (scales, veins, and surface) of this area.

Strengths:

The data and figures support the conclusions. The article is swiftly written and makes an interesting evolutionary comparison to the function of this gene in Drosophila. Based on the data presented, it can now be established that mirror likely has a similar selector function for posterior-wing identity in a plethora of insects.

We thank the reviewer for their feedback.

Weaknesses:

This first version has minor terminological issues regarding the use of the terms "domains" and "compartment".

We acknowledge that the terminologies “domains” and “compartments” might lead to confusion. To avoid confusion we have removed the term “compartment” from the manuscript.

Reviewer #2 (Public Review):

This is a short and unpretentious paper. It is an interesting area and therefore, although much of this area of research was pioneered in flies, extending basic findings to butterflies would be worthwhile. Indeed, there is an intriguing observation but it is technically flawed and these flaws are serious.

The authors show that mirror is expressed at the back of the wing in butterflies (as in flies). They present some evidence that is required for the proper development of the back of the wing in butterflies (a region dubbed the vannus by the ancient guru Snodgrass). But there are problems with that evidence. First, concerning the method, using CRISP they treat embryos and the expectation is that the mirror gene will be damaged in groups of cell lineages, giving a mosaic animal in which some lines of cells are normal for mirror and others are not. We do not know where the clones or patches of cells that are defective for mirror are because they are not marked. Also, we do not know what part of the wing is wild type and what part is mutant for mirror. When the mirror mutant cells colonise the back of the wing and that butterfly survives (many butterflies fail to develop), the back of the wing is altered in some selected butterflies. This raises a second problem: we do not know whether the rear of the wing is missing or transformed. From the images, the appearance of the back of the wing is clearly different from the wild type, but is that due to transformation or not? And then I believe we need to know specifically what the difference is between the rear of the wing and the main part. What we see is a silvery look at the back that is not present in the main part, is it the structure of the scales? We are not told.

Thank you for this feedback. We appreciate that many readers may not accustomed to looking at mosaic knockouts. As discussed in a previous review article (Zhang & Reed 2017), we rely on a combination of contralateral asymmetry and replicates to infer mutant phenotypes. For many genes (e.g. pigmentation enzymes) mutant clones are obvious, but for other types of genes (e.g. ligands) clone boundaries are sometimes not directly diagnosable. It is simply a limitation of our study system. Nonetheless, you see for yourself that “the back of the wing is altered in some butterflies” – the effects of deleting mirror are clear and repeatable.

In terms of interpreting mutant phenotypes, we agree that that paper would benefit from a better description of the specific effects. Therefore, we have included an improved, more systematic description of phenotypes, along with better-annotated figures showing changes in wing shape and venation, scale coloration, and color pattern transformation (e.g. posterior elongation of the orange marginal stripes).

There are other problems. Mirror is only part of a group of genes in flies and in flies both iroquois and mirror are needed to make the back of the wing, the alula (Kehl et al). What is known about iro expression in butterflies?

In Drosophila mirror, araucan, and caupolican comprise the so-called Iroqouis Complex of genes. As denoted in Figure S4 and in Kerner et al (doi: https://doi.org/10.1186/1471-2148-9-74) the divergence of araucan and caupolican into two separate paralogs is restricted to Drosophila. As in most insects, butterflies have only two Iroquois Complex genes: araucan and mirror. We tested the role of araucan in Junonia coenia as shown in our pre-print: https://doi.org/10.1101/2023.11.21.568172. Its expression appears to be restricted to early pupal wings where it is transcribed in all scale-forming cells. Mosaic araucan KOs resulted in a change in scale iridescent coloration associated with changes in the laminar thickness of scale cells.

In flies, mirror regulates a late and local expression of dpp that seems to be responsible for making the alula. What happens in butterflies? Would a study of the expression of Dpp in wildtype and mirror compromised wings be useful?

We thank the reviewer for the proposal and agree that a future study comparing Dpp in wild-type versus mirror KO butterflies would be useful to clarify the mechanism of Dpp signalling in wing development. It is not clear, however, that the results of a Dpp experiment would change the conclusions of our current study therefore we decided not to undertake these additional experiments for our revision.

Thus, I find the paper to be disappointing for a general journal as it does little more than claim what was discovered in Drosophila is at least partly true in butterflies.

We respect that the reviewer does not have a strong interest in the comparative aspects of this study. Fair enough. This report is primarily aimed at biologists interested in the evolutionary history of insect wings.

Also, it fails to explain what the authors mean by "wing domains" and "domain specification". They are not alone, butterfly workers, in general, appear vague about these concepts, their vagueness allowing too much loose thinking.

A domain is “a region distinctively marked by some physical feature”. This term is used extensively in the developmental biology literature (e.g. “expression domain”, “embryonic domain”, “tissue domain”, “domain specification”) and is found throughout popular textbooks (e.g. Alberts et al. “The Cell”, Gilbert “Developmental Biology”). We prefer the term “domain” because of its association in the Drosophila literature with transcription factors that define fields of cells. We specifically avoided using the term “compartment” because of its association with cell lineage, which we have not tested.

Since these matters are at the heart of the purpose and meaning of the work reported here, we readers need a paper containing more critical thought and information. I would like to have a better and more logical introduction and discussion.

We would like the very same thing, of course, and we hope the reviewer finds our revised manuscript to be more satisfying to read.

The authors do define what they mean by the vannus of the wing. In flies the definition of compartments is clear and abundantly demonstrated, with gene expression and requirement being limited precisely to sets of cells that display lineage boundaries. It is true that domains of gene expression in flies, for example of the iroquois complex, which includes mirror, can only be related to patterns with difficulty. Some recap of what is known plus the opinion of the authors on how they interpret papers on possible lineage domains in butterflies might also be useful as the reader, is no wiser about what the authors might mean at the end of it!

We thank the reviewer for this suggestion. However, our experiments have little to contribute to the topic of cell lineage compartmentalization. We have therefore opted to avoid speculating on this topic to prevent confusion and to keep the manuscript focused on our experimental results.

The references are sometimes inappropriate. The discovery of the AP compartments should not be referred to Guillen et al 1995, but to Morata and Lawrence 1975. Proofreading is required.

We thank the reviewer for suggesting this important reference. We have included it in our revision.

Reviewer #3 (Public Review):

Summary:

The manuscript by Chatterjee et al. examines the role of the mirror locus in patterning butterfly wings. The authors examine the pattern of mirror expression in the common buckeye butterfly, Junonia coenia, and then employ CRISPR mutagenesis to generate mosaic butterflies carrying clones of mirror mutant cells. They find that mirror is expressed in a well-defined posterior sector of final-instar wing discs from both hindwings and forewings and that CRISPR-injected larvae display a loss of adult wing structures presumably derived from the mirror expressing region of hindwing primordium (the case for forewings is a bit less clear since the mirror domain is narrower than in the hindwing, but there also do seem to be some anomalies in posterior regions of forewings in adults derived from CRISPR injected larvae). The authors conclude that the wings of these butterflies have at least three different fundamental wing compartments, the mirror domain, a posterior domain defined by engrailed expression, and an anterior domain expressing neither mirror nor engrailed. They speculate that this most posterior compartment has been reduced to a rudiment in Drosophila and thus has not been adequately recognized as such a primary regional specialization.

Critique:

This is a very straightforward study and the experimental results presented support the key claims that mirror is expressed in a restricted posterior section of the wing primordium and that mosaic wings from CRISPR-injected larvae display loss of adult wing structures presumably derived from cells expressing mirror (or at least nearby). The major issue I have with this paper is the strong interpretation of these findings that lead the authors to conclude that mirror is acting as a high-level gene akin to engrailed in defining a separate extreme posterior wing compartment. To place this claim in context, it is important in my view to consider what is known about engrailed, for which there is ample evidence to support the claim that this gene does play a very ancestral and conserved function in defining posterior compartments of all body segments (including the wing) across arthropods.

(1) Engrailed is expressed in a broad posterior domain with a sharp anterior border in all segments of virtually all arthropods examined (broad use of a very good panspecies anti-En antibody makes this case very strong).

(2) In Drosophila, marked clones of wing cells (generated during larval stages) strictly obey a straight anterior-posterior border indicating that cells in these two domains do not normally intermix, thus, supporting the claim that a clear A/P lineage compartment exists.

In my opinion, mirror does not seem to be in the same category of regulator as engrailed for the following reasons:

(1) There is no evidence that I am aware of, either from the current experiments, or others that the mirror expression domain corresponds to a clonal lineage compartment. It is also unclear from the data shown in this study whether engrailed is co-expressed with mirror in the posterior-most cells of J. coenia wing discs. If so, it does not seem justified to infer that mirror acts as an independent determinant of the region of the wing where it is expressed.

(2) Mirror is not only expressed in a posterior region of the wing in flies but also in the ventral region of the eye. In Drosophila, mirror mutants not only lack the alula (derived approximately from cells where mirror is expressed), but also lack tissue derived from the ventral region of the eye disc (although this ventral tissue loss phenotype may extend beyond the cells expressing mirror).

In summary, it seems most reasonable to me to think of mirror as a transcription factor that provides important development information for a diverse set of cells in which it can be expressed (posterior wing cells and ventral eye cells) but not that it acts as a high-level regulator as engrailed.

Recommendation:

While the data provided in this succinct study are solid and interesting, it is not clear to me that these findings support the major claim that mirror defines an extreme posterior compartment akin to that specified by engrailed. Minimally, the authors should address the points outlined above in their discussion section and greatly tone down their conclusion regarding mirror being a conserved selector-like gene dedicated to establishing posterior-most fates of the wing. They also should cite and discuss the original study in Drosophila describing the mirror expression pattern in the embryo and eye and the corresponding eye phenotype of mirror mutants: McNeill et al., Genes & Dev. 1997. 11: 1073-1082; doi:10.1101/gad.11.8.1073.

We thank the reviewer for their summary, critique, and recommendations. We agree with everything the reviewer says. Honestly, however, we were surprised by these comments because we took great care in the paper to never refer to mirror as a compartmentalization gene or claim it has a function in cell lineage compartmentalization like engrailed. As pointed out, we lack clonal analyses to test for compartmentalization. This is why we used the term “domain” instead of “compartment” in the title and throughout the manuscript. Nevertheless, we have recrafted the discussion in the manuscript, including completely removing the term “compartment”, to better avoid implications that mirror plays a role in cell lineage compartmentalization.

We also thank the reviewer for recommending the paper about the role of mirror in eye development. For the sake of keeping the paper focused, however, we decided not to broach the topic of mirror functions outside the context of wing development.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

I have minor comments for improvement.

The abstract and introductions are terminologically problematic when they refer to the concept of compartment and compartment boundaries. Allegedly this confusion has previously propagated in several articles related to butterfly wing development, which keeps alienating this literature from being taken seriously by fly specialists, for example. So it is important to use the right terms. I will try to explain point by point here, but I would appreciate it if the authors could undertake a significant rewrite taking these comments into account. The authors use the terms compartment and compartment boundary. This has a very specific use in developmental genetics: mitotic clones never cross a boundary (or compartment). I think the authors can keep referring to the equivalent of the A-P boundary, which is situated somewhere between M1-M2 based on unpublished data from the Patel Lab, and is not very well defined (Engrailed expression moves a little bit during development in this area). Domain is a looser term and can be used more liberally to describe genetically defined regions.

- "Classical morphological work subdivides insect wings into several distinct domains along the antero-posterior (AP) axis, each of which can evolve relatively independently." Yes. This concept of domain and individuation seems important. You could make a proposed link to selector genes here.

- "There has been little molecular evidence, however, for AP subdivision beyond a single compartment boundary described from Drosophila melanogaster." Incorrect, and this conflates "domain" and "compartment".

Flies have wing AP domains too, that pattern their veins (see the cited Banerjee et al).

- "Our results confirm that insect wings can have more than one posterior developmental domain, and support models of how selector genes may facilitate evolutionarily individuation of distinct AP domains in insect wings". Yes, and I like the second part of the sentence. Still, I would recommend simply deleting "confirm that insect wings can have more than one posterior developmental domain, and" because this is neglecting previous work on AP genetic regionalization in both flies (vein literature) and butterflies (e.g. McKenna and Nijhout, Banerjee et al).

- "Analyses of wing pattern diversity across butterflies, considering both natural variation and genetic mutants, suggest that wings can be subdivided into at least five AP domains, bounded by the M1, M3, Cu2, and 2A veins respectively, within each of which there are strong correlations in color pattern variation and wing morphology (Figure 1A)". Yes, and I would recommend emphasizing they correspond to welldefined gene expression domains as mentioned in Banerjee et al, or McKenna and Nijhout.

- "The anterior-most of these domains, bordered by the M1 vein, appears to correspond to an AP compartment boundary originally described by cell lineage tracing in Drosophila melanogaster, and later supported in butterfly wings by expression of the Engrailed transcription factor. Interestingly, however, D. melanogaster work has yet to reveal clear evidence for additional AP domain boundaries in the wing." Confusingly, because the first sentence is about compartments while the second is about AP domains. I also think the claim that Dmel has no other known AP domains is dubious because Spalt is highly regionalized in flies.

- "Previous authors have proposed the existence of such individuated domains, and speculated that they may be specified by selector genes.5,10 Our data provide experimental support for this model, and now motivate us to identify factors that specify other domain boundaries between the M1 and A2 veins." Yes, I completely agree with this way to emphasize the selector effect, and to link it to the concept of "individuated domain"

We cannot thank the reviewer enough for the time and thought they devoted to giving helpful suggestions to improve our manuscript. We have applied all of the above recommendations to the revision.

Fig. S1: the field needs to move away from Red/Green microscopy images, for accessibility reasons.

The easiest fix here would be to change the red channels to magenta.

Green/Magenta provides excellent contrast and accessibility in general in 2-channel images.

We thank the reviewer for this suggestion. We have improved the color accessibility of Fig. S1.

Version 2

Abstract

Abstract

Insect wings, a key innovation that contributed to the explosive diversification of insects, are recognized for their remarkable variation and many splendid adaptations. Classical morphological work subdivides insect wings into several distinct domains along the antero-posterior (AP) axis, each of which can evolve relatively independently to produce the myriad forms we see in nature. Important insights into AP subdivision of insect wings comes from work in Drosophila melanogaster, however they do not fully explain the diversity of AP domains observed across broad winged insects. Here we show that the transcription factor mirror acts as a selector gene to differentiate a far posterior domain in the butterfly wing, classically defined as the vannus, and has effects on wing shape, scale morphology, and color pattern. Our results support models of how selector genes may facilitate evolutionarily individuation of distinct AP domains in insect wings outside of Drosophila, and suggest that the D. melanogaster wing blade has been reduced to represent only a portion of the archetypal insect wing.

Results and discussion

Insect wings vary dramatically in their size and morphology. Butterflies and moths in particular have attracted attention for their rapidly evolving wings that show extreme variation in shape and color pattern. Interestingly, the diversity of lepidopteran color patterns appears to follow relatively simple rules, wherein an evolutionarily conserved set spot and stripe pattern elements can vary in size, color, and presence/absence along the antero-posterior (AP) axis.1,2 Surveys of wing pattern diversity across butterflies, considering both natural variation and genetic mutants, suggest that wings can be subdivided into five AP domains, bounded by the M1, M3, Cu2, and 2A veins as summarized in Figure 1A.3 Within each domain there is a strong correlation between pattern variation and wing morphology, and, conversely, between domains there is a relative degree of independence in morphological variation. We know little about how these different AP domains are differentiated during wing development, however.

Proposed AP domains of insect wings.

(A) Comparative morphology previously identified boundaries associated with color pattern variation, demarcated by the M1, M3, Cu2, and 2A veins.5 (B) Examples of the 2A vein marking a clear posterior color pattern boundary across multiple butterfly families. (C) Proximal posterior features of the archetypal dipteran wing include the calypters, which are only found in Calypterata clade, and the alula, a lobe at the base of the wing blade.21 (D) mirror RNAi knockdowns result in highly specific loss of the alula in D. melanogaster.13 (E) Snodgrass’ model of the archetypal insect wing specifies three major domains along the anteroposterior axis: the remigium, the vannus, and the jugum. Homologies of these domains with the features of the dipteran wing (C), or butterfly wing pattern boundaries (A), has remained unclear.

Some aspects of the AP wing differentiation process have been extrapolated from D. melanogaster wings to butterflies. The transcription factors engrailed and spalt define AP domains in developing fly wings4-6 and exhibit expression patterns in butterflies that suggest some functions may be conserved. In butterflies engrailed transcription across the entire region posterior to the M1 vein, while its paralog invected marks a largely overlapping region between M1 and 2A.7-9 spalt expression is more dynamic and complex in butterflies, but appears to consistently mark a region between the R2 and M3 veins.9 These observations, coupled with the overall size and morphological complexity of butterfly wings, have spurred interest in the question of whether additional AP differentiation mechanisms may underlie AP domain specification in Lepidoptera and other insects.10,11 The identification of additional domain-determining transcription factors in broad winged insects would provide a model for the modular diversification of insect wing morphology.

The goal of this study was to identify the molecular basis of AP domain specification in the butterfly wing, posterior of the M1 engrailed boundary. We initially identified the transcription factor mirror as a potential candidate selector gene based on previous mRNA-seq studies,12 coupled with D. melanogaster work showing that mirror specifies the alula – a small membranous lobe at the posterior base of the wing in some dipterans (Figure 1C,D).13,14 We used in situ hybridization to visualize mirror mRNA localization in the last-instar imaginal discs of the common buckeye butterfly, Junonia coenia, and observed mirror expression throughout the wing domain posterior to the 2A vein boundary (Figures 2A and S1). This expression precisely marks the vannus, or anal region (Figures 1B, E), of the butterfly wing – a distinct posterior domain of the wing blade populated by the anal (A) veins and bordered anteriorly by the claval fold between the cubital (Cu) and A veins in the hindwing (Figure 2B). As observed in various representative species across different butterfly families, this region of the hindwing is distinct from the rest of the wing; the color patterns defining the wing blade terminate at the 2A vein, and in many species it is marked by scales that differ in morphology and lack color patterns found in anterior regions of the wing (Figures 1B and S2A). This region has been proposed to represent the posterior-most color pattern domain in butterflies (Figure 1A, Figure 2B).3

mirror determines the identity of the far posterior vannus domain in J. coenia.

(A) In J. coenia mirror (red) is expressed posterior to the 2A vein in the late last-instar hindwing imaginal discs, marking the region that will develop into the vannus. See Figure S1 for forewing expression and controls. (B) In adult butterfly wings the vannus is the wing field posterior to the 2A vein, shown here in J. coenia as a posterior lobe with silvery scales devoid of color patterns found anterior to 2A. (C) Targeted mosaic knockouts (mKOs) of mirror result in loss of vannus features in J. coenia. (D) Annotation of wild-type features preserved on one side of a contralateral mKO. (E) Annotation of full effect mutant phenotyp in the opposing wing from the individual in (D) highlights loss of lobe, loss of 3A, ectopic branching of 2A, and posterior extension of color patterns. (F) Annotation of a partial mKO highlights transformation of posterior lobe and extension of color patterns, both of which terminate at the mutant clonal boundary. (G) The individual shown in (F) also shows ectopic branching of 2A in the mutant region, but preservation of 3A in the wild-type region. For additional mirror mKO phenotypes see Figures S2-S4.

To functionally assess the role of mirror in wing development we generated CRISPR/Cas9 mosaic knockouts (mKOs) targeting the coding region of J. coenia mirror. We recovered many mKO mutants that consistently exhibited several distinctive phenotypes (Table S1), but only ever in the mirror-expressing posterior vannus region:

Venation: In mKO mutant hindwings we observed frequent truncation or complete loss of the loss of the 3A vein, as well as discontinuity and ectopic bifurcation of 2A veins (hindwings: Figures 2C-G, S2, S3; forewings: Figure S4).

Wing shape: mKOs consistently showed shape change along the posterior edge of the wing, especially the hindwing where the projecting lobe characteristic of the vannus was reduced and replaced by a more even, regular wing margin. The anal fold between the 2A and 3A veins, which marks the edge of the vannal lobe, was also lost in mutant clones affecting this wing region (Figures 2C-G, S2-S3).

Color pattern expansion: In mutant hindwings we consistently observed extension of wing margin color patterns across the 2A vein boundary into the posterior region of the wing. In wild-type wings the submarginal and central symmetry system color patterns terminate at the 2A vein (dorsal: Figure 2C-G; ventral: Figure S3). Remarkably, in mirror mKO mutants these entire color pattern systems extend past the 2A vein into the posterior regions of the wing on both dorsal and ventral wing surfaces. Specifically, submarginal bands continue past 2A to follow the posterior wing margin (Figures 2C-G, S2), while central symmetry patterns extend from 2A down to the posterior margin (Figure S3).

Scale morphology: In most mutants vannus scales lost their distinctive silver structural color and rounded apical edges, and transformed into brown, serrated scales typical of the remigium (Figure S2A). We also observed that scales along mutant wing margins often displayed elongated scales, often manifesting as a dense hair-like matt across the posterior margin (Figure S2A).

The knockout phenotypes we observed illustrate several things about mirror’s function in wing development. First, across all mKO replicates we never observed mutant phenotypes anterior to the 2A vein, which suggests that mirror effects are limited to the far posterior domain in which we observed mirror expression. Second, mirror knockouts appear to affect multiple aspects of wing morphology, including scale shape and coloration, overall wing shape, and wing venation. Finally, the loss of mirror function resulted in transformation, not loss, of the posterior domain. Even in mKO individuals that appeared to have a mutant phenotype across the entire posterior region, the mirror-expressing cells posterior to the 2A vein were not simply lost – the wing field continues well past the 2A vein. Instead, the identities of the post-2A scale cells changed in color and structure. In this respect, the expansion of multiple color pattern systems into the posterior domain of mutants was particularly compelling because the loss of central symmetry system and submarginal patterns in the post-2A vannus region is commonly observed in Lepidoptera (Figure 1B). The ability to reconstitute these complex color pattern elements in the posterior domain by knocking out a single transcription factor leads us to speculate that their patterning information may occur in a latent form that is masked by mirror effects. Together, our expression and knockout data suggest that mirror acts as a selector gene necessary to define the identity of the far posterior wing domain.

Identifying mirror as a vannus-specifying factor is of interest for a few reasons. First, it provides a molecular explanation for how different AP domains of the butterfly wing may be individuated to independently evolve their own shape and color pattern variations. Previous authors have proposed the existence of such individuated domains, and speculated that they may be specified by selector genes.3,10 Our data provide experimental support for this model, and now motivate us to identify additional factors that may specify other domain boundaries between the M1 and A2 veins.

Next, the role of mirror in specifying the vannus provides an important link between molecular developmental biology and Snodgrass’ classical anatomical designations of the insect wing fields – i.e., the remigium, the vannus, and the jugum (Figure 1E).15 The remigium and the vannus are dominant features of most insect wings, while the jugum is smaller lobe-like feature that is reduced or lost in many clades. The remigium is the major anterior wing blade that is directly articulated by thoracic motor musculature to power flight, and is typically divided from the vannus by a distinct wing fold (e.g. the “anal fold” in butterflies). In many insect orders the vannus is a prominent fan-like structure, which can be a greatly enlarged gliding surface some hemimetabolous clades such as Orthoptera (crickets and grasshoppers), Dictyoptera (mantids and roaches), Phasmatodea (stick insects), and Plecoptera (stoneflies). Importantly, mirror knockdown in the hemimetaboulous milkweed bug Oncopeltus fasciatus causes developmental defects in the claval and anal furrows in the posterior wing,16 which leads us to infer that mirror’s role in determining the vannus may be deeply conserved in insects. This prompts us to speculate that the developmental individuation of the posterior domain by mirror facilitated the remarkable diversification of the vannus across the insects.

Finally, the function of mirror in determining the alula in D. melanogaster13 suggests that the alula may represent an evolutionarily reduced vannus. The ramifications of this are significant for reconstructing the history of the insect wing, because they suggest that the D. melanogaster wing blade is a lone remigium – only one-third of the archetypal insect wing (Figure 1E). The dipteran jugum appears to have been lost entirely, except perhaps as a membranous haltere cover (calypter) in the Calyptratae clade, as speculated by Snodgrass.15 Thus, while flies have been an important model for characterizing genes and processes that build insect wings, a more comprehensive understanding of the development and evolution of insect wings requires work in species that have wings more representative of the complete ancestral blueprint.